---
title: "Cyclistics"
author: "HuynhTan"
date: "`r Sys.Date()`"
output: html_document
---

## Using Library

helps wrangle data

Use the conflicted package to manage conflicts

```{r}
library(tidyverse) 
library(conflicted)
```

```{r}
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

# STEP 1: COLLECT DATA
```{r}
m_012024 <- read.csv("202401-divvy-tripdata.csv")
m_022024 <- read.csv("202402-divvy-tripdata.csv")
m_032024 <- read.csv("202403-divvy-tripdata.csv")
m_042024 <- read.csv("202404-divvy-tripdata.csv")
m_052024 <- read.csv("202405-divvy-tripdata.csv")
m_062024 <- read.csv("202406-divvy-tripdata.csv")
```

## View data
```{r}
head(m_012024)
#head(m_022024)
#head(m_032024)
#head(m_042024)
#head(m_052024)
#head(m_062024)
```

## Inspect the dataframes and look for incongruencies
```{r}
str(m_012024)
#str(m_022024)
# str(m_032024)
# str(m_042024)
# str(m_052024)
```
# STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE

## Stack individual month's data frames into one big data frame
```{r}
all_trips <- bind_rows(m_012024, m_022024, m_032024, m_042024, m_052024, m_062024)
```

```{r}
str(all_trips)
```
## Remove lat, long
```{r}
all_trips <- all_trips %>% 
  select(-c(start_lat, start_lng, end_lat, end_lng))
``` 

# STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS

## Inspect the new table

```{r}
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(all_trips)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics
```
## Check to make sure the proper number of observations were reassigned

```{r}
table(all_trips$member_casual)
```
## Add columns that list the date, month, day, year and ride_length of each ride
```{r}
all_trips$started_at <- as.POSIXct(all_trips$started_at, format="%Y-%m-%d %H:%M:%S")
all_trips$ended_at <- as.POSIXct(all_trips$ended_at, format="%Y-%m-%d %H:%M:%S")
```

```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

## Inspect the structure of the columns
```{r}
str(all_trips)
```

## Remove "bad" data

```{r}
all_trips_v2 <- all_trips[!(all_trips$ride_length<0), ]
```

```{r}
head(all_trips_v2)
str(all_trips_v2)
```
# STEP 4: CONDUCT DESCRIPTIVE ANALYSIS
## Descriptive analysis on ride_length (all figures in seconds)

```{r}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride
```

```{r}
summary(all_trips_v2$ride_length)
```
## Compare members and casual users

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```
##See the average ride time by each day for members vs casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```
# analyze ridership data by type and weekday
```{r}
all_trips_v2 %>%
  mutate(weekday = wday(started_at, label = TRUE)) %>% #creates weekday field using wday()
  group_by(member_casual, weekday) %>% #groups by usertype and weekday
  summarise(number_of_rides = n() #calculates the number of rides
            ,average_duration = mean(ride_length)) %>% # calculates the average duration
  arrange(member_casual, weekday) # sorts
```
# Visualize the number of rides by rider type

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

# Create a visualization for average duration

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

```
# STEP 5: EXPORT SUMMARY FILE FOR FURTHER ANALYSIS

```{r}
results <- all_trips_v2 %>%
  group_by(member_casual, day_of_week) %>%
  summarise(
    number_of_rides = n(),
    average_duration = mean(ride_length)
  ) %>%
  arrange(member_casual, day_of_week)  # Sắp xếp theo member_casual và day_of_week

write.csv(results, file = 'avg_ride_length_and_counts.csv', row.names = FALSE)
```

